---
title: "DEseq analysis of Gat201-dependent gene expression"
author: "Liz Hughes and Edward Wallace"
date: "20/12/2021"
output:
  html_document:
    toc: yes
    toc_depth: 2
  word_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , warning=FALSE, message=FALSE)
```

# Summary

This document is an analysis of the Gat201 RNA-seq dataset, focusing on differences that depend only on Gat201 at later timepoints.

The first goal is heatmap of Gat201-dependent differential gene expression.

A later goal is lists of Gat201-DEGs for comparison with ChIP-seq.

# Load packages used for analysis

```{r load_packages, include=FALSE}

library(tidyr)
library(readr)
library(dplyr)
library(readxl)
library(stringr)
library(ggplot2)
library(cowplot)
library(biobroom)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(magrittr)
library(markdown)
library(forcats)
library(genefilter)
library(ggrepel)


theme_set(
  theme_cowplot(font_size = 12) +
    theme(panel.grid.major = element_line(colour = "grey80", size = 0.5))
)

```


# Load count data and remove unwanted parts of column names

```{r load_count_data}
# note: we should insist that counts are loaded as integers
# and check the other data types
counts <- readr::read_tsv("../quantseqfwd_EH_050221/counts.txt",
                   comment = "#") %>% 
          dplyr::rename_with(str_remove_all, pattern = "_S[0-9]+_R1_001_aln.bam")

```


# Load sample sheet and format for DESeq2's requirements: readxl::read_excel

```{r load_sample_sheet}

Condition_Time_levels <- c("Y_0", 
                           "R_30", "R_120", "R_240",
                           "RS_30", "RS_120", "RS_240") 

samplesheet <- readxl::read_excel("../input_experiment/Gat201_samplesheet.xlsx") %>%
    magrittr::set_rownames(.$SampleID) %>%
    dplyr::mutate(GAT201 = 
                    forcats::fct_collapse(Strain, 
                                          "WT" = c("a","A"), 
                                          "del" = c("B","M")) %>% 
                    forcats::fct_relevel("del"),
                  Timepoint = factor(Timepoint),
                  Condition_Time = factor(
                    paste(Condition, Time, sep = "_"),
                    levels = Condition_Time_levels
                  )
    )

```

# Select counts

```{r select_counts_all}

counts_all <-
      dplyr::select(counts, samplesheet$SampleID) %>%
      magrittr::set_rownames(counts$Geneid)

```

# Construct the data object from the matrix of counts and the samplesheet information table

```{r deseq_Condition_Time_GAT201}

dds_all <- DESeqDataSetFromMatrix(countData = counts_all,
                                   colData = samplesheet,
                                   design = ~ Condition_Time + GAT201:Condition_Time - 1 ) %>%
  DESeq()

dds_all
resultsNames(dds_all)
```

```{r tidy_deseq_Condition_Time_GAT201}
coefnames_GAT201 <- c("Condition_TimeY_0.GAT201WT", "Condition_TimeR_30.GAT201WT",
                      "Condition_TimeR_120.GAT201WT",  "Condition_TimeR_240.GAT201WT",
                      "Condition_TimeRS_30.GAT201WT",  "Condition_TimeRS_120.GAT201WT",
                      "Condition_TimeRS_240.GAT201WT")

# Create a tidy data frame that contains only the GAT201-dependent log2FC
# in a helpful format for plotting
deseq_df_CT_vsGAT201 <- 
  biobroom::tidy.DESeqDataSet(dds_all) %>%
  dplyr::filter(term %in% coefnames_GAT201) %>%
  mutate(Condition_Time = term %>%
           stringr::str_remove("Condition_Time") %>%
           stringr::str_remove(".GAT201WT") %>%
           factor(levels = Condition_Time_levels)) %>%
  select(Condition_Time, gene, baseMean, log2FC = estimate, stderror, padj = p.adjusted)

# check we retained all the levels
unique(deseq_df_CT_vsGAT201$Condition_Time)
```

### Reality-check by plotting fold change for a couple of known genes

Check that some marker genes are behaving as expected:

- GAT204 CNAG_06762, up in WT
- TSA1 CNAG_06917, up in WT
- ACT1 CNAG_00483, not really affected

```{r plotpoints_log2FC_select}

ggplot( data = dplyr::filter(deseq_df_CT_vsGAT201, 
                             gene %in% c("CNAG_06762", "CNAG_06917", "CNAG_00483"))
        ) +
  geom_point(aes(x = Condition_Time, y = log2FC, colour = gene, shape = gene))
  
```

```{r plotheat_log2FC_select}

plotheat_log2FC <- function( genes = c("CNAG_06762", "CNAG_06917", "CNAG_00483"),
                             data = deseq_df_CT_vsGAT201) {
  ggplot( data = dplyr::filter(data, 
                               gene %in% genes)
  ) +
    geom_tile(aes(x = Condition_Time, y = gene, fill = log2FC)) +
    scale_fill_gradient2("log2 fold-change\nWT vs ∆GAT201",
                         high = "darkblue", low = "darkred") +
    coord_cartesian(expand = c(0,0)) +
    theme(panel.grid.major = element_blank(),
          axis.line = element_blank(),
          axis.ticks.y = element_blank())
}

plotheat_log2FC()

plotheat_log2FC(genes = c("CNAG_05392", "CNAG_06246", "CNAG_03438"))
```

### Select genes that are DE dependent on GAT201 at any timepoint

Here we also plot in order of log2FC at R_240.

```{r select_DEGAT201}
DEGAT201_up_anytime <- 
  deseq_df_CT_vsGAT201 %>%
  dplyr::filter(log2FC > 1, padj < 0.05) 

DEGAT201_down_anytime <- 
  deseq_df_CT_vsGAT201 %>%
  dplyr::filter(log2FC < -1, padj < 0.05)

DEGAT201_list_anytime <- c(unique(DEGAT201_up_anytime$gene),
                           unique(DEGAT201_down_anytime$gene))

# list genes arranged by log2FC at R_240
DEGAT201_list_anytime_arrange_R_240 <- 
  deseq_df_CT_vsGAT201 %>%
  dplyr::filter(gene %in% DEGAT201_list_anytime, Condition_Time == "R_240") %>%
  dplyr::arrange(log2FC) %>%
  dplyr::pull(gene)


plotheat_log2FC(genes = DEGAT201_list_anytime, 
                  data = deseq_df_CT_vsGAT201 %>%
                    dplyr::filter(gene %in% DEGAT201_list_anytime) %>%
                    dplyr::mutate(gene = factor(gene, 
                                                levels = DEGAT201_list_anytime_arrange_R_240))
  ) + 
  theme(axis.text.y = element_blank())
```

# To do

- do a better ordering or clustering of the heatmap
- fix x-axis labels on heatmap to be informative
- maybe make a smaller plot with sample gene names printed?
