---
title: "DEseq analysis of Gat201-dependent gene expression"
author: "Edward Wallace"
date: "2023-04-24"
output:
  html_document:
    toc: yes
    toc_depth: 2
  word_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: '2'
---


# Summary

This document is an analysis of the Gat201 RNA-seq dataset, comparing only 0min with 2h RPMI + Serum for wild-type. The goal is to compare the differential gene expression in this dataset with matched conditions in the wakeup dataset, despite the very different RNA-seq protocols.


# Setup: load packages, load data, clean and prep data


## Load packages used for analysis

```{r load_packages, include=FALSE}

library(tidyr)
library(readr)
library(dplyr)
library(readxl)
library(stringr)
library(ggplot2)
library(cowplot)
library(biobroom)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(magrittr)
library(markdown)
library(forcats)
library(genefilter)
library(ggrepel)


theme_set(
  theme_cowplot(font_size = 12) +
    theme(panel.grid.major = element_line(colour = "grey80", size = 0.5))
)

```


## Load count data and remove unwanted parts of column names

```{r load_count_data}
# note: we should insist that counts are loaded as integers
# and check the other data types
counts <- readr::read_tsv("../quantseqfwd_EH_050221/counts.txt",
                   comment = "#") %>% 
          dplyr::rename_with(str_remove_all, pattern = "_S[0-9]+_R1_001_aln.bam")

```


## Load sample sheet and format for DESeq2's requirements: readxl::read_excel

We pick the condition `RS_120`, RPMI + Serum at 2hrs, along with `Y_0`, YPD at 0mins.

```{r load_sample_sheet}

Condition_Time_levels <- c("Y_0", 
                           "R_30", "R_120", "R_240",
                           "RS_30", "RS_120", "RS_240") 

samplesheet <- readxl::read_excel("../input_experiment/Gat201_samplesheet.xlsx") %>%
    magrittr::set_rownames(.$SampleID) %>%
    dplyr::mutate(GAT201 = 
                    forcats::fct_collapse(Strain, 
                                          "WT" = c("a","A"), 
                                          "del" = c("B","M")) %>% 
                    forcats::fct_relevel("del"),
                  Timepoint = factor(Timepoint),
                  Condition_Time = factor(
                    paste(Condition, Time, sep = "_"),
                    levels = Condition_Time_levels
                  )
    )

samplesheet_WTRS2h_vs_0min <- samplesheet %>%
  dplyr::filter(Condition_Time %in% c("Y_0", "RS_120"), GAT201 == "WT")
```

## Load H99 Gene information

```{r load_gene_info}
gene_info_df <- 
  here::here("input_annotation",
             "H99_geneinfo_FungiDB59.txt") %>%
  readr::read_tsv(comment = "#")  %>% 
  dplyr::rename(gene = GeneID, geneName = GeneName)
```

# Differential gene expression analysis with DESEq2

## Run DESeq2 analysis dependent on Condition/Time and GAT201

This first uses the matrix of counts and the samplesheet information table to construct the DESeq2 data object, then immediately runs the analysis.


```{r deseq_WTRS2h_vs_0min}

dds_WTRS2h_vs_0min <- DESeqDataSetFromMatrix(
    countData = dplyr::select(counts, samplesheet_WTRS2h_vs_0min$SampleID) %>%
      magrittr::set_rownames(counts$Geneid),
    colData = samplesheet_WTRS2h_vs_0min,
    design = ~ Timepoint ) %>%
  DESeq()

dds_WTRS2h_vs_0min
```


```{r results_WTRS2h_vs_0min}
results(dds_WTRS2h_vs_0min)

resultsNames(dds_WTRS2h_vs_0min)
```

```{r deseq_df_WTRS2h_vs_0min}
deseq_df_WTRS2h_vs_0min <- 
  biobroom::tidy.DESeqDataSet(dds_WTRS2h_vs_0min) %>%
  dplyr::filter(term == "Timepoint_120_vs_0") %>%
  dplyr::left_join(gene_info_df,
                   by = "gene") %>% 
  dplyr::select(gene, 
                geneName, 
                baseMean, 
                log2FC = estimate, 
                stderror,
                padj = p.adjusted)
```

## List DEGs most UP in RPMI+ vs YPD

We cut off at 2x changes with 5% FDR here.


```{r DEGs_RPMI_up, fig.width = 6, fig.height = 3}
# Make data frames of DEGs up and down, arranged by log2FC
DEGdf_RPMI_up2x_FDR5 <- 
  dplyr::filter(deseq_df_WTRS2h_vs_0min,
                log2FC > 1, padj < 0.05) %>%
  dplyr::arrange(desc(log2FC))
print(DEGdf_RPMI_up2x_FDR5, n = 20)
```

Print the top 20 *named* DEGs

```{r DEGs_RPMI_up_named}
DEGdf_RPMI_up2x_FDR5 %>%
  dplyr::filter(!is.na(geneName)) %>%
  print(n = 20)
```

## List DEGs most DOWN in RPMI+ vs YPD


```{r DEGs_RPMI_down, fig.width = 6, fig.height = 3}
# Make data frames of DEGs up and down, arranged by log2FC
DEGdf_RPMI_down2x_FDR5 <- 
  dplyr::filter(deseq_df_WTRS2h_vs_0min,
                log2FC < -1, padj < 0.05) %>%
  dplyr::arrange(log2FC)
print(DEGdf_RPMI_down2x_FDR5, n = 20)
```

Print the top 20 *named* DEGs

```{r DEGs_RPMI_down_named}
DEGdf_RPMI_down2x_FDR5 %>%
  dplyr::filter(!is.na(geneName)) %>%
  print(n = 20)
```

## Volcano plot of DEGs

```{r volcano_WTRS2h_vs_0min, fig.width = 6, fig.height = 3}
volcano_WTRS2h_vs_0min <- 
  ggplot(data = deseq_df_WTRS2h_vs_0min,
         aes(x = log2FC, y = -log10(padj)) ) +
  geom_point(size = 0.5, colour = "grey50") +
  geom_point(data = DEGdf_RPMI_up2x_FDR5,
             size = 1, colour = "darkblue") +
  geom_point(data = DEGdf_RPMI_down2x_FDR5,
             size = 1, colour = "darkred") +
  scale_y_continuous("-log10(p)",
                     limits = c(0,10), expand = c(0,0),
                     oob=scales::squish) +
  scale_x_continuous("log2 fold-change, RPMI+serum vs YPD, 37C 1hr",
                     breaks = -5:5,
                     limits = c(-4,4), expand = c(0.01,0.01),
                     oob = scales::squish) +
  coord_cartesian(clip = "off") 
volcano_WTRS2h_vs_0min
```



## Write outputs to files

```{r write_DEGs}
deseq_df_WTRS2h_vs_0min %>%
  dplyr::mutate(
    across(c(baseMean, log2FC, stderror), round, digits = 3)
  ) %>%
  dplyr::mutate(padj = signif(padj, digits = 4)) %>%
  write_tsv(file = here::here("results",
                              "deseq_df_WTRS2h_vs_0min.txt"))

# DEGdf_RPMI_up2x_FDR5 %>%
#   dplyr::pull(gene) %>%
#   write_lines(file = here::here("results", "DEGs",
#                             "deglist_WTRS2h_vs_0min_up2x_FDR5.txt"))
# 
# DEGdf_RPMI_down2x_FDR5 %>%
#   dplyr::pull(gene) %>%
#   write_lines(file = here::here("results", "DEGs",
#                             "deglist_WTRS2h_vs_0min_down2x_FDR5.txt"))
```

## Compare differential expression in Gat201 dataset with Wakeup dataset

```{r load_wakeup}
deseqdf_wakeup <- 
  read_tsv("../../CryptoWakeupRNASeq/results/DEGs/deseq_df_RPMI37C2hr_vs_0min.txt",
           comment = "#")

deseq_joined <- 
  dplyr::inner_join(
    deseqdf_wakeup,
    deseq_df_WTRS2h_vs_0min,
    by = c("gene", "geneName"),
    suffix = c("_wakeup", "_Gat201")
)

ggplot(data = deseq_joined) + 
  geom_density(aes(x = baseMean_wakeup, colour = "Wakeup"))  + 
  geom_density(aes(x = baseMean_Gat201, colour = "Gat201")) +
  scale_x_log10("baseMean")

deseq_joined
```

Correlation is overall low:

```{r calc_correlation}
deseq_joined %>%
  with(cor(log2FC_wakeup, log2FC_Gat201,
      use = "pairwise.complete.obs"))
```

Correlation is still low if we filter for low-expressed genes by removing bottom 10% of baseMean:

```{r calc_correlation_filtered_q25}
cor_log2FC_q10filtered <- 
  deseq_joined %>%
  dplyr::filter(baseMean_wakeup > quantile(baseMean_wakeup, 0.1), 
                baseMean_Gat201 > quantile(baseMean_wakeup, 0.1)) %>%
  with(cor(log2FC_wakeup, log2FC_Gat201,
           use = "pairwise.complete.obs"))

cor_log2FC_q10filtered
```

Scatter plot for filtered data show that the dynamic range of Gat201 dataset is greater than the Wakeup dataset:

```{r plot_scatter_filtered, fig.height = 5.4, fig.width = 5}
plot_scatter_filtered <- 
ggplot(data = deseq_joined %>%
  dplyr::filter(baseMean_wakeup > quantile(baseMean_wakeup, 0.1), 
                baseMean_Gat201 > quantile(baseMean_wakeup, 0.1)),
       aes(x = log2FC_wakeup, y = log2FC_Gat201)) +
  geom_vline(xintercept = 0, size = 0.5, colour = "grey80") +
  geom_hline(yintercept = 0, size = 0.5, colour = "grey80") +
  geom_point(size = 0.5, colour = "grey50") +
  scale_x_continuous("log2 fold-change, Wakeup dataset",
                     breaks = -6:6,
                     limits = c(-7,7), expand = c(0.01,0.01),
                     oob = scales::squish) +
  scale_y_continuous("log2 fold-change, Gat201 dataset",
                     breaks = -6:6,
                     limits = c(-7,7), expand = c(0.01,0.01),
                     oob = scales::squish) +
  coord_equal(clip = "off") +
  labs(title = "RPMI+serum vs YPD, 37°C 2hr, wild-type") +
  theme(panel.grid.major = element_blank())

plot_scatter_filtered
```

```{r plot_scatter_filtered_selectgenes, fig.height = 6.4, fig.width = 6}
my30genes <- 
  read_tsv("../input_annotation/CryptoWakeup_30genes.txt") %>%
  dplyr::rename(gene = GeneID)

deseq_joined_my30genes <- 
  left_join(my30genes, 
            deseq_joined %>% dplyr::select(-geneName),
            by = "gene")
  
plot_scatter_filtered +
  annotate("text", x = 7, y = 7, hjust = 1, vjust = 1,
           label = paste0("R = ", round(cor_log2FC_q10filtered,2))) +
  geom_point(data = deseq_joined_my30genes, size = 1, colour = "darkblue") +
  geom_text_repel(data = deseq_joined_my30genes, 
                  aes(label = GeneName), colour = "darkblue")
```

