---
title: "PCA analysis of Gat201 RNA-seq dataset"
author: "Liz Hughes and Edward Wallace"
date: "16/12/2021"
output:
  html_document:
    toc: yes
    toc_depth: 2
  word_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , warning=FALSE, message=FALSE)
```

# Summary

This document is a big-picture analysis of trends in the Gat201 RNA-seq dataset.

The first goal is a principal component analysis (PCA) to show the main axes of variation in the data.

# Load packages used for analysis

```{r load_packages, include=FALSE}

library(tidyr)
library(readr)
library(dplyr)
library(readxl)
library(stringr)
library(ggplot2)
library(cowplot)
library(biobroom)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(magrittr)
library(markdown)
library(forcats)
library(genefilter)
library(ggrepel)


theme_set(
  theme_cowplot(font_size = 12) +
    theme(panel.grid.major = element_line(colour = "grey80", size = 0.5))
)

```


# Load count data and remove unwanted parts of column names

```{r load_count_data}

counts <- readr::read_tsv("../quantseqfwd_EH_050221/counts.txt",
                   comment = "#") %>% 
          dplyr::rename_with(str_remove_all, pattern = "_S[0-9]+_R1_001_aln.bam")

```


# Load sample sheet and format for DESeq2's requirements: readxl::read_excel

```{r load_sample_sheet}

samplesheet <- readxl::read_excel("../input_experiment/Gat201_samplesheet.xlsx") %>%
    magrittr::set_rownames(.$SampleID)%>%
    dplyr::mutate(GAT201 = forcats::fct_collapse(Strain, WT = c("a","A"), Gat201 = c("B","M")))

```

# Select counts

```{r select_counts_all}

counts_all <-
      dplyr::select(counts, samplesheet$SampleID) %>%
      magrittr::set_rownames(counts$Geneid)

```

# Construct the data object from the matrix of counts and the samplesheet information table

```{r dds_counts_all}

(dds_all <- DESeqDataSetFromMatrix(countData = counts_all,
                                   colData = samplesheet,
                                   design = ~ Condition  + Time + GAT201))

```


# rlog transformation

```{r rlog_dds_all}

rld_all <- rlog(dds_all)
           head(assay(rld_all))
```


# PCA Plot function for PC1 and PC2

```{r function_12}

plotPCA <- function (object, intgroup = "condition", ntop = 500, returnData = FALSE) 
{
    rv <- rowVars(assay(object))
    select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, 
        length(rv)))]
    pca <- prcomp(t(assay(object)[select, ]))
    percentVar <- pca$sdev^2/sum(pca$sdev^2)
    if (!all(intgroup %in% names(colData(object)))) {
        stop("the argument 'intgroup' should specify columns of colData(dds)")
    }
    intgroup.df <- as.data.frame(colData(object)[, intgroup, 
        drop = FALSE])
    group <- if (length(intgroup) > 1) {
        factor(apply(intgroup.df, 1, paste, collapse = " : "))
    }
    else {
        colData(object)[[intgroup]]
    }
    d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], PC3 = pca$x[, 3], PC4 = pca$x[, 4], PC5 = pca$x[, 5], group = group, 
        intgroup.df, name = colnames(object))
    if (returnData) {
        attr(d, "percentVar") <- percentVar[1:2]
        return(d)
    }
    ggplot(data = d, aes_string(x = "PC1", y = "PC2", color = "group")) + 
        geom_point(size = 3) + xlab(paste0("PC1: ", round(percentVar[1] * 
        100), "% variance")) + ylab(paste0("PC2: ", round(percentVar[2] * 
        100), "% variance")) + coord_fixed()
    
  }

```

# PCA_1-2

```{r plot_PC12}

plotPCA(rld_all, intgroup = c("GAT201", "Time"))


```

