---
title: "PCA-2 Analysis"
author: "Liz Hughes"
date: "07/12/2021"
output: html_document
---
---
title: "PCA Plots_2"
author: "Liz Hughes"
date: "07/12/2021"
output:
  html_document:
    toc: yes
    toc_depth: 2
  word_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , warning=FALSE, message=FALSE)
```

# Load packages used for analysis
```{r Load packages used for analysis, include=FALSE}

library(tidyr)
library(readr)
library(dplyr)
library(readxl)
library(stringr)
library(ggplot2)
library(cowplot)
library(biobroom)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(magrittr)
library(markdown)
library(forcats)
library(genefilter)
library(ggrepel)


theme_set(
  theme_cowplot(font_size = 12) +
    theme(panel.grid.major = element_line(colour = "grey80", size = 0.5))
)

```


# Load count data and remove unwanted parts of column names

```{r Load count data}

counts <- readr::read_tsv("~/GitHub/CryptoGat201RNASeq_draft_b/quantseqfwd_EH_050221/counts.txt",
                   comment = "#") %>% 
          dplyr::rename_with(str_remove_all, pattern = "_S[0-9]+_R1_001_aln.bam")

```

# Load sample sheet and format for DESeq2's requirements: readxl::read_excel

```{r Load sample sheet}

samplesheet <- readxl::read_excel("~/GitHub/CryptoGat201RNASeq_draft_b/input_experiment/Gat201_samplesheet.xlsx") %>%
    magrittr::set_rownames(.$SampleID)%>%
    dplyr::mutate(GAT201 = forcats::fct_collapse(Strain, WT = c("a","A"), Gat201 = c("B","M")))

```


# Select counts

```{r Select counts_all}

counts_all <-
  dplyr::select(counts, samplesheet$SampleID) %>%
   magrittr::set_rownames(counts$Geneid)

```

# Construct the data object from the matrix of counts and the samplesheet information table

```{r dds_all_counts_all}

(dds_all <- DESeqDataSetFromMatrix(countData = counts_all,
                                  colData = samplesheet,
                                  design = ~ Condition  + Time + GAT201))

```


# PCA Plot for All using rlog transformation

## PC1-2

```{r function_12}

#getMethod("plotPCA","DESeqTransform")
#DESeq2:::plotPCA.DESeqTransform

plotPCA <- function (object, intgroup = "condition", ntop = 500, returnData = FALSE) 
{
    rv <- rowVars(assay(object))
    select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, 
        length(rv)))]
    pca <- prcomp(t(assay(object)[select, ]))
    percentVar <- pca$sdev^2/sum(pca$sdev^2)
    if (!all(intgroup %in% names(colData(object)))) {
        stop("the argument 'intgroup' should specify columns of colData(dds)")
    }
    intgroup.df <- as.data.frame(colData(object)[, intgroup, 
        drop = FALSE])
    group <- if (length(intgroup) > 1) {
        factor(apply(intgroup.df, 1, paste, collapse = " : "))
    }
    else {
        colData(object)[[intgroup]]
    }
    d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], PC3 = pca$x[, 3], PC4 = pca$x[, 4], PC5 = pca$x[, 5], group = group, 
        intgroup.df, name = colnames(object))
    if (returnData) {
        attr(d, "percentVar") <- percentVar[1:2]
        return(d)
    }
    ggplot(data = d, aes_string(x = "PC1", y = "PC2", color = "group")) + 
        geom_point(size = 3) + xlab(paste0("PC1: ", round(percentVar[1] * 
        100), "% variance")) + ylab(paste0("PC2: ", round(percentVar[2] * 
        100), "% variance")) + coord_fixed()
    
  }

```

## Print PCA_12 Time and then Condition

```{r Plot_PC12}
print(plotPCA(rld_all, intgroup = c("GAT201", "Time")))
print(plotPCA(rld_all, intgroup = c("GAT201", "Condition")))


```


## PC2-3
```{r function_23}

plotPCA <- function (object, intgroup = "condition", ntop = 500, returnData = FALSE) 
{
    rv <- rowVars(assay(object))
    select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, 
        length(rv)))]
    pca <- prcomp(t(assay(object)[select, ]))
    percentVar <- pca$sdev^2/sum(pca$sdev^2)
    if (!all(intgroup %in% names(colData(object)))) {
        stop("the argument 'intgroup' should specify columns of colData(dds)")
    }
    intgroup.df <- as.data.frame(colData(object)[, intgroup, 
        drop = FALSE])
    group <- if (length(intgroup) > 1) {
        factor(apply(intgroup.df, 1, paste, collapse = " : "))
    }
    else {
        colData(object)[[intgroup]]
    }
    d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], PC3 = pca$x[, 3], PC4 = pca$x[, 4], PC5 = pca$x[, 5], group = group, 
        intgroup.df, name = colnames(object))
    if (returnData) {
        attr(d, "percentVar") <- percentVar[2:3]
        return(d)
    }
    ggplot(data = d, aes_string(x = "PC2", y = "PC3", color = "group")) + 
        geom_point(size = 3) + xlab(paste0("PC2: ", round(percentVar[2] * 
        100), "% variance")) + ylab(paste0("PC3: ", round(percentVar[3] * 
        100), "% variance")) + coord_fixed()
    
  }

```
## Print PCA_23 Time and then Condition

```{r Plot_PC23}
print(plotPCA(rld_all, intgroup = c("GAT201", "Time")))
print(plotPCA(rld_all, intgroup = c("GAT201", "Condition")))


```


## PC3-4
```{r function_34}

plotPCA <- function (object, intgroup = "condition", ntop = 500, returnData = FALSE) 
{
    rv <- rowVars(assay(object))
    select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, 
        length(rv)))]
    pca <- prcomp(t(assay(object)[select, ]))
    percentVar <- pca$sdev^2/sum(pca$sdev^2)
    if (!all(intgroup %in% names(colData(object)))) {
        stop("the argument 'intgroup' should specify columns of colData(dds)")
    }
    intgroup.df <- as.data.frame(colData(object)[, intgroup, 
        drop = FALSE])
    group <- if (length(intgroup) > 1) {
        factor(apply(intgroup.df, 1, paste, collapse = " : "))
    }
    else {
        colData(object)[[intgroup]]
    }
    d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], PC3 = pca$x[, 3], PC4 = pca$x[, 4], PC5 = pca$x[, 5], group = group, 
        intgroup.df, name = colnames(object))
    if (returnData) {
        attr(d, "percentVar") <- percentVar[3:4]
        return(d)
    }
    ggplot(data = d, aes_string(x = "PC3", y = "PC4", color = "group")) + 
        geom_point(size = 3) + xlab(paste0("PC3: ", round(percentVar[3] * 
        100), "% variance")) + ylab(paste0("PC4: ", round(percentVar[4] * 
        100), "% variance")) + coord_fixed()
    
  }

```

## Print PCA_34 Time and then Condition

```{r Plot_PC34}
print(plotPCA(rld_all, intgroup = c("GAT201", "Time")))
print(plotPCA(rld_all, intgroup = c("GAT201", "Condition")))


```


## PC4-5
```{r function_45}

plotPCA <- function (object, intgroup = "condition", ntop = 500, returnData = FALSE) 
{
    rv <- rowVars(assay(object))
    select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, 
        length(rv)))]
    pca <- prcomp(t(assay(object)[select, ]))
    percentVar <- pca$sdev^2/sum(pca$sdev^2)
    if (!all(intgroup %in% names(colData(object)))) {
        stop("the argument 'intgroup' should specify columns of colData(dds)")
    }
    intgroup.df <- as.data.frame(colData(object)[, intgroup, 
        drop = FALSE])
    group <- if (length(intgroup) > 1) {
        factor(apply(intgroup.df, 1, paste, collapse = " : "))
    }
    else {
        colData(object)[[intgroup]]
    }
    d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], PC3 = pca$x[, 3], PC4 = pca$x[, 4], PC5 = pca$x[, 5], group = group, 
        intgroup.df, name = colnames(object))
    if (returnData) {
        attr(d, "percentVar") <- percentVar[4:5]
        return(d)
    }
    ggplot(data = d, aes_string(x = "PC4", y = "PC5", color = "group")) + 
        geom_point(size = 3) + xlab(paste0("PC4: ", round(percentVar[4] * 
        100), "% variance")) + ylab(paste0("PC5: ", round(percentVar[5] * 
        100), "% variance")) + coord_fixed()
    
  }

```

## Print PCA_45 Time and then Condition

```{r Plot_PC45}
print(plotPCA(rld_all, intgroup = c("GAT201", "Time")))
print(plotPCA(rld_all, intgroup = c("GAT201", "Condition")))


```
