---
title: "Serum does not affect gene expression much"
author: "Liz Hughes and Edward Wallace"
date: "12/01/2022"
output:
  html_document:
    toc: yes
    toc_depth: 2
  word_document:
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , warning=FALSE, message=FALSE)
```

# Summary

This document is an analysis of the Gat201 RNA-seq dataset, focusing on differences that depend on serum. This means that we treat the two WT strains as the same, and the two ∆gat201 strains as the same.

This makes figure panels:

- heatmap of log2FC for all (or most) genes, showing overall variability, strong dependence on time, weak dependence on serum.

These will then be combined into a multipanel figure, planned as figure 3 of the manuscript.

# Setup: load packages, load data, clean and prep data


## Load packages used for analysis

```{r load_packages, include=FALSE}

library(tidyr)
library(readr)
library(dplyr)
library(readxl)
library(stringr)
library(ggplot2)
library(cowplot)
library(biobroom)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(magrittr)
library(markdown)
library(forcats)
library(genefilter)
library(ggrepel)


theme_set(
  theme_cowplot(font_size = 12) +
    theme(panel.grid.major = element_line(colour = "grey80", size = 0.5))
)

```


## Load count data and remove unwanted parts of column names

```{r load_count_data}
# note: we should insist that counts are loaded as integers
# and check the other data types
counts <- readr::read_tsv("../quantseqfwd_EH_050221/counts.txt",
                   comment = "#") %>% 
          dplyr::rename_with(str_remove_all, pattern = "_S[0-9]+_R1_001_aln.bam")

```


## Load sample sheet and format for DESeq2's requirements: readxl::read_excel

Here we choose how the Strain labels behave by manipulating the factors. This creates a new variable `GAT201` whose levels are `del` and `WT`.

```{r load_sample_sheet}

Condition_Time_levels <- c("Y_0", 
                           "R_30", "R_120", "R_240",
                           "RS_30", "RS_120", "RS_240")

# GAT201_Condition_Time
GCT_levels <- c("WT_Y_0", 
                "WT_R_30", "WT_RS_30", 
                "WT_R_120", "WT_RS_120", 
                "WT_R_240", "WT_RS_240",
                "del_Y_0", 
                "del_R_30", "del_RS_30", 
                "del_R_120", "del_RS_120", 
                "del_R_240", "del_RS_240")

samplesheet <- readxl::read_excel("../input_experiment/Gat201_samplesheet.xlsx") %>%
    magrittr::set_rownames(.$SampleID) %>%
    dplyr::mutate(GAT201 = 
                    forcats::fct_collapse(Strain, 
                                          "WT" = c("a","A"), 
                                          "del" = c("B","M")) %>% 
                    forcats::fct_relevel("del"),
                  Timepoint = factor(Timepoint),
                  Condition_Time = factor(
                    paste(Condition, Time, sep = "_"),
                    levels = Condition_Time_levels
                  ),
                  GCT = factor(
                    paste(GAT201, Condition, Time, sep = "_"),
                    levels = GCT_levels
                  )
    )

```

## Select counts

```{r select_counts_all, warning = FALSE}

counts_all <-
      dplyr::select(counts, samplesheet$SampleID) %>%
      magrittr::set_rownames(counts$Geneid)

```

# Differential gene expression analysis with DESeq2

## Run DESeq2 analysis dependent oncombination GAT201, Condition, and Time

This first uses the matrix of counts and the samplesheet information table to construct the DESeq2 data object, then immediately runs the analysis.

The design formula `~ GCT - 1` ensures that every combination of GAT201, Condition and Time has a coefficient, where we remove the intercept term by including `- 1`. Removing the intercept term means that we have a coefficient for the first value of GCT, i.e. `WT_Y_0`.

```{r deseq_GAT201_Condition_Time}

dds_all <- DESeqDataSetFromMatrix(
    countData = counts_all,
    colData = samplesheet,
    design = ~ GCT - 1 ) %>%
  DESeq()

dds_all
resultsNames(dds_all)
```


## Select GAT201-dependent log2FC at each timepoint

```{r tidy_deseq_GCT}
# Create a tidy data frame that contains only the GAT201-dependent log2FC
# in a helpful format for plotting
deseq_df_GCT <- 
  biobroom::tidy.DESeqDataSet(dds_all) %>%
  mutate(GCT = term %>%
           stringr::str_remove("GCT") %>%
           factor(levels = GCT_levels)) %>%
  select(GCT, gene, baseMean, log2estimate = estimate) %>%
  group_by(gene) %>%
  mutate(log2FC = log2estimate - mean(log2estimate)) %>%
  ungroup()
  

# check we retained all the levels
unique(deseq_df_GCT$GCT)
```

## Sanity-check: Did we get the log2FC right fora  favourite gene?

```{r print_log2FC_6917}
deseq_df_GCT %>%
  filter(gene == "CNAG_06917")
```

```{r plot_log2FC_6917}
ggplot(data = deseq_df_GCT %>%
  filter(gene %in% c("CNAG_00483", "CNAG_00121", "CNAG_06917", "CNAG_01551")),
  aes(x = GCT, fill = log2FC, y = gene)) +
  geom_tile() +
  scale_fill_gradient2(low = "cyan", mid = "black", high = "yellow") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.1),
        panel.grid.major = element_blank(),
        axis.line = element_blank())
```
## Plot log2FC for all genes, with no clustering

```{r plot_log2FC_all}
ggplot(data = deseq_df_GCT,
  aes(x = GCT, fill = log2FC, y = gene)) +
  geom_tile() +
  scale_fill_gradient2(low = "cyan", mid = "black", high = "yellow",
                       limits = c(-3.5, 3.5), oob = scales::squish) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0.1),
        panel.grid.major = element_blank(),
        axis.line = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank())
```
## Calculate gene expression hierarchical clusters

This does not run properly, perhaps because there are NA/NaN/Inf values?

It does run if we restrict to a subset with `head()`.

```{r hclust_genes, eval = FALSE}
log2FC_GCT_wide <-
  deseq_df_GCT %>%
  select(gene,GCT, log2FC) %>%
  pivot_wider(id_cols = gene, names_from = GCT, values_from = log2FC) %>%
  select(c("gene", GCT_levels))

hclust_log2FC <- 
  log2FC_GCT_wide %>%
  set_rownames(.$gene) %>%
  select(-gene) %>%
  head(n= 100) %>%
  dist() %>%
  hclust()
```

# TO DO

- Check where the zero values are in the dataset
- Filter for higher-expressed genes if needed?
- Debug the calls to dist() and hclust() on larger dataset
- Include gene order from `hclust_log2FC` alongside `deseq_df_GCT`
- Plot the heatmap with the clustered gene order
- If possible, plot the dendrogram alongside the heatmap
